Synthesizing Quotient Lenses

### 2017-11-14 papers/QuotientLenses/main.tex
make build succeed on first building
added proof.sty and some comments for Solomon

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  % BCP: I find this distracting:
  % stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

@@

\section{Introduction}
{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.

@@

@@

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are
difficult to express, or that cannot be expressed, using the other
combinators.

However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences which occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

@@

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\noindent The behaviour of \cd{ambiguous} is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a", "aa", "aaa"} depending on how
\cd{"aaa"} is parsed. We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.

To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

@@

\begin{figure}[p!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\project{R}{s}$ \textbf{WF}}
\DisplayProof
\hskip 1.5em 
\end{center}

@@

The most complicated inferrence rule is the rule for the \textit{permutation}
combinator:

@@

As we mentioned before, each QRE $c$ enables us to express:
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole language
$W(c)$ of a QRE. The \textit{squash} and \textit{permutation} combinators
give the two interesting definitions . If $c = \squash{c}{c'}{f}$, then
the whole language of $c$ is $W(c) \sep W(c')$. This is because the
\textit{squash} combinator takes the whole language $W(c)$ of $c$, the whole
language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c)) \longrightarrow
\mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into $\mathcal{L}(W(c'))$ using
$f$ and then canonizes $W(c')$ into $K(c')$ using $\canonizer(c')$.

@@

 \begin{figure}[ht]
  \centering
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$\kw{lift}(\ell): id(R) \Leftrightarrow id(S)$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}
  \end{align*}

@@

In our previous work, we showed that given regular expressions $R, S$ and a set
of input-output pairs $\{(r_1, s_1), \ldots, (r_n, s_n)\}$, if there is a
derivation $\ell : R \Leftrightarrow S$ of a bijective lens that maps $r_i$ to
$s_i$ in the forward direction and $s_i$ to $r_i$ in the backward direction,
then there is an algorithm that computes a bijective lens $\ell' : R'
\Leftrightarrow S'$ such that $R \equiv R'$, $S \equiv S'$, $\ell'$ maps $r_i$
to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward direction,
and $\llbracket \ell \rrbracket = \llbracket \ell' \rrbracket$~\cite{popl18}.

@@

Our evaluations serve to answer the following questions:
\begin{enumerate}
\item Does synthesizing quotient lenses from QREs permit for an easier
  development process than existing approaches.

\item Are QREs sufficiently expressive that more lenses can be synthesized
  without modifications to the format.

\item Is our synthesis algorithm competitive with existing lens synthesis
  tools.
\end{enumerate}

@@

Optician was only able to synthesize fully bijective data transformations.
\Name loosens that restriction, and is able to synthesize data which is
bijective when put in canonical form.
To determine the increase of expressivity, we analyzed:
\begin{enumerate}
\item How many data formats in Optician's benchmark suite required modifications
  to be bijective.
\item How many data formats in Optician's benchmark suite required modifications
  to be bijective when put in canonical form.
\end{enumerate}

@@

\bibliography{local}

>>>>

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  % BCP: I find this distracting:
  % stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

% For Inference Rules
\newcommand{\Rule}[2]{\infer{#2}{#1}}
\newcommand{\RuleSide}[3]{\infer[#3]{#2}{#1}}
%\newcommand{\Axiom}[1]{\Rule{}{#1}}

\newcommand{\wf}[1]{\vdash \ensuremath{#1\;\mathsf{wf}}}

@@

\section{Introduction}

{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.

@@

\saz{Fix the notation in the following code.}

@@

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. \saz{rewrite the next sentence to
  avoid so many uses of ``express''?} The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are
difficult to express, or that cannot be expressed, using the other
combinators.

\saz{Add something about being ``complete'' here?  Say more about how normalize
  wouldn't be used in ``ordinary'' QRE programming?}
However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences which\saz{``which'' $\to$ ``that''} occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

@@

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity}, which comes up\saz{``comes up'' is a little colloquial; perhaps ``arises''?} when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\noindent The behaviour of \cd{ambiguous} is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"} depending on how
\cd{"aaa"} is parsed.

\saz{Add citations to prior work that uses these definitions?}
We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.
%
To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

@@

\begin{figure}[p!]
\saz{This well-formedness judgment is better, but the \textbf{WF} is a bit
  jarring -- it detracts from the content of the judgment.  Maybe something
  lighter, like just $\mathit{wf}$?}
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\project{R}{s}$ \textbf{WF}}
\DisplayProof
\hskip 1.5em 
\end{center}

@@

The most complicated inferrence rule is the rule for the \textit{permutation}
combinator: \saz{fix the overflow by stacking ?  Also, it should use the WF
  predicate in the premise, right?}

@@

As we mentioned before, each QRE $c$ enables us to express: \saz{I'm not sure we
  need to repeat the whole list here.}
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole language
$W(c)$ of a QRE. The \textit{squash} and \textit{permutation} combinators
give the two interesting definitions. If $c = \squash{c}{c'}{f}$, then
the whole language of $c$ is $W(c) \sep W(c')$. This is because the
\textit{squash} combinator takes the whole language $W(c)$ of $c$, the whole
language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c)) \longrightarrow
\mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into $\mathcal{L}(W(c'))$ using
$f$ and then canonizes $W(c')$ into $K(c')$ using $\canonizer(c')$.

@@

 \begin{figure}[ht]
  \centering

\[
\begin{array}[b]{c@{\qquad}c}
\Rule{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow id(S)}
&
  \begin{array}[b]{rcl}
  \llbracket q \rrbracket.get &=&  \llbracket \ell \rrbracket\\
  \llbracket q \rrbracket.put &=& \llbracket \ell \rrbracket^{-1}
  \end{array}
\end{array}
\]

% \begin{tabular}[t]{cc}

% \begin{minipage}{\textwidth/2}
%   \begin{prooftree}
% \AxiomC{$\ell : R \Leftrightarrow S$}
% \UnaryInfC{$\kw{lift}(\ell): id(R) \Leftrightarrow id(S)$}
% \end{prooftree}
% \end{minipage}
% &
% \(
% \)
% \end{tabular}

\[
\Rule{q' : c'  \Leftrightarrow c'' &
 \wf{c} &
 K(c) = W(c')}
{\kw{lquot}(c, q'): c' \circ c \Leftrightarrow c''}
\]

@@

In our previous work \saz{citation}, we showed that given regular expressions $R, S$ and a set
of input-output pairs $\{(r_1, s_1), \ldots, (r_n, s_n)\}$, if there is a
derivation $\ell : R \Leftrightarrow S$ of a bijective lens that maps $r_i$ to
$s_i$ in the forward direction and $s_i$ to $r_i$ in the backward direction,
then there is an algorithm that computes a bijective lens $\ell' : R'
\Leftrightarrow S'$ such that $R \equiv R'$, $S \equiv S'$, $\ell'$ maps $r_i$
to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward direction,
and $\llbracket \ell \rrbracket = \llbracket \ell' \rrbracket$~\cite{popl18}.

@@

Our evaluations serve to answer the following questions:
\begin{enumerate}
\item Does synthesizing quotient lenses from QREs permit for an easier
  development process than existing approaches?

\item Are QREs sufficiently expressive that more lenses can be synthesized
  without modifications to the format?

\item Is our synthesis algorithm competitive with existing lens synthesis
  tools?
\end{enumerate}

@@

Optician was only able to synthesize fully bijective data transformations.
\Name{} loosens that restriction, and is able to synthesize data which is
bijective when put in canonical form.
To determine the increase of expressivity, we analyzed:
\begin{enumerate}
\item How many data formats in Optician's benchmark suite required modifications
  to be bijective.
\item How many data formats in Optician's benchmark suite required modifications
  to be bijective when put in canonical form.
\end{enumerate}

@@

\bibliographystyle{plain}
\bibliography{local}
